Sampling Clint and Funbound.plasma

### 2025-08-22 data-raw/ClearanceProject-0-Distinguishability.Rmd
Updates project file organization. Moves common functions to a common, sourced file.

```{r library_loading}
# June 6, 2025 8AM: Using httk-dev/HTTK-293 branch
# June 6, 2025 3PM: Using httk-dev/dev branch
source("data-raw/RC_comparison_bkgd.R")
library(dplyr)
library(tidyselect)
library(ggplot2)
library(patchwork)
library(parallel)
library(mirai)
library(ctxR)
```

@@

```{r set_parameters}
THIS_CHEMICAL = "DTXSID3024289" # A chemical to test
DOSE = 25
SPECIES = "Rat"
ROUTE = "oral"
TIMES = c(seq(0, 23.75, by = 0.25), seq(24, 72, by = 6),
          seq(96, 120, by = 24))/24
SAMPLES = 1000
VARPARS = list(Clint = 0.1)
CENSORPARS = list(Funbound.plasma = list(cv = 0.1, lod = 1E-5))
HUMAN_CLINT_FUP = TRUE
SPECIES_TEXT <- ifelse(SPECIES != "Human" && HUMAN_CLINT_FUP,
                       paste0(SPECIES, "_parHuman"),
                       SPECIES)
INVITROUV = NULL

@@

```{r helper_functions}
# Get all simulations (restrictive and nonrestrictive)
conc_fun_wsamples <- function(
    # Parameterizing args
    dtxsid,
    species = "Human",
    default_to_human = FALSE,
    force_human = FALSE,
    # Solving args
    dose = 20, 
    route = "iv",
    times = seq(0, 48, by = 1)/24,
    # MC sampler args
    vary.params = NULL,
    censored.params = NULL,
    samples = 1000
) {
  # Decided to do both Clearances in one go
  # parameterization to be able to explicitly force human clint and fup

  # Parameterize
  gas_pars <- standard_parameterization(
    dtxsid = dtxsid,
    species = species,
    default_to_human = default_to_human,
    force_human = force_human
  )

  # SOLVE NUMERICALLY
  numerical_solution <- standard_httk_preds(parameters = gas_pars,
                                            species = species,
                                            default_to_human = default_to_human,
                                            force_human = force_human,
                                            dose = dose,
                                            route = route,
                                            times = times)

  mc_solutions <- standard_httk_mc(parameters = gas_pars,
                                   species = species,
                                   default_to_human = default_to_human,
                                   force_human = force_human,
                                   dose = dose,
                                   route = route,
                                   times = times,
                                   vary.params = vary.params,
                                   censored.params = censored.params,
                                   samples = samples)

  # Merge the solutions
  tmp <- merge(numerical_solution, mc_solutions)
  tmp <- subset(tmp, Time %in% times)

  # Time from days to hours
  tmp$Time <- janitor::round_to_fraction(tmp$Time * 24, denominator = 60)

  return(tmp)
}

@@

for (this_chem in TEST_CHEMICALS) {
  solved_mc_fun <- conc_fun_wsamples(this_chem,
                                     species = SPECIES,
                                     default_to_human = FALSE,
                                     force_human = HUMAN_CLINT_FUP,
                                     vary.params = VARPARS,
                                     censored.params = CENSORPARS,
                                     dose = DOSE,
                                     route = ROUTE,
                                     times = TIMES)
  print(
    ggplot(solved_mc_fun, aes(x = Time)) +
      geom_line(mapping = aes(y = MED_R), color = "magenta4",
                linewidth = 1, linetype = "76") +
      geom_ribbon(mapping = aes( ymin = l95_R, ymax = u95_R), linetype = "dotted",
                  fill = NA, color = "magenta4") +
      geom_line(mapping = aes(y = MED_N), color = "cyan4",
                linewidth = 1, linetype = "76") +
      geom_ribbon(mapping = aes(ymin = l95_N, ymax = u95_N), linetype = "dotted",
                  fill = NA, color = "cyan4") +
      geom_line(aes(y = Sol_N), linetype = "55") +
      geom_line(aes(y = Sol_R), linetype = "solid") +
      labs(y = "Concentration (mg/L)",x = "Time (hrs)",
           title = paste0(this_chem, " (", DOSE, "mg/kg)"),
           subtitle = paste0("Samples: ", SAMPLES)) +
      theme_bw()
  )
  Sys.sleep(10)
}

#### Set-up assessment functions

We need to assess the distinguishability in a standard manner, with a few 
considerations. We can assess human and rat at arbitrary doses and coefficients
of variation for `Clint` and `Funbound.plasma`. However, there may be specific
doses and CVs for which we want to test. Note that there are

```{r all_chems_fun}

assessment_fun <- function(
    x,
    species = SPECIES,
    dose = DOSE,
    route = ROUTE,
    times = TIMES,
    samples = SAMPLES,
    vary.params = VARPARS,
    censored.params = CENSORPARS,
    force_human = HUMAN_CLINT_FUP
) {
  p_df <- conc_fun_wsamples(
    dtxsid = x,
    species = species,
    dose = dose,
    route = route,
    times = times,
    samples = samples,
    vary.params = vary.params,
    censored.params = censored.params,
    force_human = force_human
  ) |> try()

@@

}

```{r parallel_sims}
# Parallelized simulation assessment
detectCores() # 16 cores, use 12
gc()

clust <- makeCluster(14L)
clusterEvalQ(clust, library("httk"))
clusterExport(clust, c("get_95CI_mc", "conc_fun_wsamples", "calc_overlap"))
clusterExport(clust, c("standard_parameterization", "standard_httk_preds",
                       "standard_httk_mc"))
clusterExport(clust, c("chem.physical_and_invitro.data"))
clusterExport(clust, c("ROUTE", "DOSE", "SAMPLES", "SPECIES", "TIMES", 
                       "VARPARS",
                       "CENSORPARS",
                       "HUMAN_CLINT_FUP"))

@@

table(sim_bins)
table(grepl("ND_SO_", sim_bins))
sum(table(sim_bins))
bindf <- data.frame(DTXSID = names(sim_bins), Bin = sim_bins,
                    CV_Clint = 0.1, CV_Funbound.plasma = 0.1, Dose = 25)
head(bindf)
rownames(bindf) <- NULL

saveRDS(bindf,
        paste0("data-raw/", SPECIES_TEXT, "_", ROUTE, INVITROUV_TEXT,
               "_simulation_bins.rds")
)

@@

```{r plot_all_chems}
# Parallelized simulation assessment
clust <- makeCluster(12L)
clusterEvalQ(clust, library("httk"))
clusterEvalQ(clust, library("ggplot2"))
clusterExport(clust, c("get_95CI_mc", "conc_fun_wsamples", "calc_overlap"))
clusterExport(clust, c("standard_parameterization", "standard_httk_preds",
                       "standard_httk_mc"))
clusterExport(clust, c("chem.physical_and_invitro.data"))
clusterExport(clust, c("ROUTE", "DOSE", "SAMPLES", "SPECIES", "TIMES", 
                       "VARPARS",
                       "CENSORPARS",
                       "HUMAN_CLINT_FUP"))

@@

pdf(file = paste0(Sys.getenv("FIG_DIR"), "RestNonRest_",SPECIES_TEXT,
                  ROUTE,"_Comparison_gasPBTK.pdf"),
    paper = "usr", bg = "white")
for (i in seq_along(sim_pls)) {
  print(sim_pls[[i]])
}
dev.off()
rm(sim_pls)
gc()

@@

```{r lombardo_sipes}

load(file = "data-raw/CYP_Sipes2017.Rdata")

@@

```{r data-specific-parameterization}
# Each x is a grouped split of a data.frame
# Using a factor of DTXSID, Species, Media, Route, Dose
invivo_prediction <- function(
    x,
    vary.params = VARPARS,
    censored.params = CENSORPARS,
    force_human = HUMAN_CLINT_FUP
) {
  dtxsid <- unique(x[["DTXSID"]])
  species <- tools::toTitleCase(unique(x[['Species']]))
  media <- unique(x[['Media']])
  route <- unique(x[['Route']])
  dose <- unique(x[['Dose']])
  times <- unique(x[['Time']])/24 # Multiple values (in hours)

  invivo_preds <- conc_fun_wsamples(
    dtxsid = dtxsid,
    species = species,
    dose = dose,
    route = route,
    times = times,
    samples = 1000,
    force_human = force_human,
    vary.params = vary.params,
    censored.params = censored.params
  ) |> try()

  if (inherits(invivo_preds, "try-error")) return(NULL)

  # If measured media is blood need to divide CPLASMA results by Rblood2plasma
  if (media == "blood") {
    Rb2p <- standard_parameterization(
      dtxsid = dtxsid, 
      species = species,
      force_human = force_human
    )[['par_rest']][['Rblood2plasma']]

    browser()
    cols2convert <- apply(
      expand.grid(c("Sol", "l95", "MED", "u95"), c("_R", "_N")),
      MARGIN = 1,
      paste0, collapse = ""
      )

    invivo_preds <- invivo_preds |>
      dplyr::mutate(dplyr::across(dplyr::all_of(cols2convert), function(x) x / Rb2p))
  }

  cbind(
    data.frame(
    DTXSID = dtxsid,
    Species = tolower(species),
    Dose = dose,
    Route = route,
    Media = media
    ),
    as.data.frame(censored.params),
    as.data.frame(vary.params),
    invivo_preds
  )
}

@@

system.time(
  temp_lst <- parLapply(clust, temp_lst, invivo_prediction)
)
stopCluster(clust)

@@

fit_tree_rat <- rpart::rpart(Bin ~ ., data = tree_rat)
rpart.plot::rpart.plot(fit_tree_rat)
preds_tree <- predict(fit_tree_rat, tree_rat, type = "class")
caret::confusionMatrix(preds_tree, tree_rat$Bin, positive = 'Distinguished')

>>>>

```{r library_loading}
source("data-raw/RC_comparison_bkgd.R")
library(dplyr)
library(tidyselect)
library(ggplot2)
library(patchwork)
library(parallel)
library(mirai)
library(ctxR)
```

@@

Here I want to understand whether selecting the restrictive or nonrestrictive
mode of clearance makes *distinguishable* differences in the predictions. This
is somewhat similar to discerning whether this parameter is structurally 
identifiable. The way I will do this is to take a sampled distribution for
Clint and Funbound.plasma then compare the 95% confidence intervals between the
estimates, labeling these parameters **Distinguishable** when there is
a *10%* (or lower) chance a sampled prediction would overlap between the
restrictive and nonrestrictive.
There are two reasons why this is quite strict threshold:
for a single bolus exposure/treatment, the differences over time collapse to
zero (or realistically, the limit of detection) and when determining distinguishability I would like to ensure that it is as clearly distinguishable as
possible.
The latter reasoning does pose a risk of false negatives (calling something
non-distinguishable in error) but this is partly assuaged by the motivation to
find a robust training set for a QSAR or QSAR-like model of clearance.
The non-distinguishable set will also play a role in the model.

```{r set_parameters}
THIS_CHEMICAL = "DTXSID3024289" # A chemical to test
DOSE = 25
SPECIES = "Rat"
ROUTE = "oral"
TIMES = c(seq(0, 23.75, by = 0.25), seq(24, 72, by = 6),
          seq(96, 120, by = 24))/24
SAMPLES = 1000
VARPARS = list(Clint = 0.1)
CENSORPARS = list(Funbound.plasma = list(cv = 0.1, lod = 1E-5))
HUMAN_CLINT_FUP = FALSE
SPECIES_TEXT <- ifelse(SPECIES != "Human" && HUMAN_CLINT_FUP,
                       paste0(SPECIES, "_parHuman"),
                       SPECIES)
INVITROUV = NULL

@@

@@

for (this_chem in TEST_CHEMICALS) {
  solved_mc_fun <- conc_fun_wsamples(this_chem,
                                     species = SPECIES,
                                     default_to_human = FALSE,
                                     force_human = HUMAN_CLINT_FUP,
                                     vary.params = VARPARS,
                                     censored.params = CENSORPARS,
                                     dose = DOSE,
                                     route = ROUTE,
                                     times = TIMES)
  print(
    ggplot(solved_mc_fun, aes(x = Time)) +
      geom_line(mapping = aes(y = MED_R), color = "magenta4",
                linewidth = 1, linetype = "76") +
      geom_ribbon(mapping = aes( ymin = l95_R, ymax = u95_R), linetype = "dotted",
                  fill = NA, color = "magenta4") +
      geom_line(mapping = aes(y = MED_N), color = "cyan4",
                linewidth = 1, linetype = "76") +
      geom_ribbon(mapping = aes(ymin = l95_N, ymax = u95_N), linetype = "dotted",
                  fill = NA, color = "cyan4") +
      geom_line(aes(y = Sol_N), linetype = "55") +
      geom_line(aes(y = Sol_R), linetype = "solid") +
      labs(y = "Concentration (mg/L)",x = "Time (hrs)",
           title = paste0(this_chem, " (", DOSE, "mg/kg)"),
           subtitle = paste0("Samples: ", SAMPLES)) +
      theme_bw()
  )
  Sys.sleep(10)
}
```

#### Assesing all chemicals

We need to assess the distinguishability in a standard manner, with a few 
considerations. We can assess human and rat at arbitrary doses and coefficients
of variation for `Clint` and `Funbound.plasma`. However, there may be specific
doses and CVs for which we want to test.

```{r helper_functions}
# Function for standardized assessment of restrictive and non-restrictive
assessment_fun <- function(
    x,
    species = SPECIES,
    dose = DOSE,
    route = ROUTE,
    times = TIMES,
    samples = SAMPLES,
    vary.params = VARPARS,
    censored.params = CENSORPARS,
    force_human = HUMAN_CLINT_FUP
) {
  p_df <- conc_fun_wsamples(
    dtxsid = x,
    species = species,
    dose = dose,
    route = route,
    times = times,
    samples = samples,
    vary.params = vary.params,
    censored.params = censored.params,
    force_human = force_human
  ) |> try()

@@

}
```

```{r parallel_sims}
# Parallelized simulation assessment
detectCores() # 16 cores, use 10
gc()

clust <- makeCluster(12L)
clusterEvalQ(clust, library("httk"))
clusterExport(clust, c("get_95CI_mc", "conc_fun_wsamples", "calc_overlap"))
clusterExport(clust, c("standard_parameterization", "standard_httk_preds",
                       "standard_httk_mc"))
clusterExport(clust, c("chem.physical_and_invitro.data"))
clusterExport(clust, c("ROUTE", "DOSE", "SAMPLES", "SPECIES", "TIMES", 
                       "VARPARS",
                       "CENSORPARS",
                       "HUMAN_CLINT_FUP"))

@@

# low, hi, mid refer to the hepatic extraction coefficient.
table(sim_bins)
table(grepl("ND_SO_", sim_bins))
sum(table(sim_bins))
bindf <- data.frame(DTXSID = names(sim_bins), Bin = sim_bins,
                    CV_Clint = 0.1, CV_Funbound.plasma = 0.1, Dose = 25)
head(bindf)
rownames(bindf) <- NULL

saveRDS(bindf,
        paste0("data-raw/RCdata/",
               SPECIES_TEXT, "_",
               ROUTE,
               INVITROUV_TEXT,
               "_simulation_bins.rds")
)

@@

```{r plot_all_chems}
# Parallelize simulation assessment
clust <- makeCluster(12L)
clusterEvalQ(clust, library("httk"))
clusterEvalQ(clust, library("ggplot2"))
clusterExport(clust, c("get_95CI_mc", "conc_fun_wsamples", "calc_overlap"))
clusterExport(clust, c("standard_parameterization", "standard_httk_preds",
                       "standard_httk_mc"))
clusterExport(clust, c("chem.physical_and_invitro.data"))
clusterExport(clust, c("ROUTE", "DOSE", "SAMPLES", "SPECIES", "TIMES", 
                       "VARPARS",
                       "CENSORPARS",
                       "HUMAN_CLINT_FUP"))

@@

pdf(file = paste0(Sys.getenv("FIG_DIR"),
                  "RestrictiveClearance_Analysis/RestNonRest_",
                  SPECIES_TEXT,
                  ROUTE,"_Comparison_gasPBTK.pdf"),
    paper = "usr", bg = "white")

for (i in cli::cli_progress_along(sim_pls)) {
  print(sim_pls[[i]])
}
dev.off()
rm(sim_pls)
rm(clust)
gc()

@@

```{r lombardo_sipes}
# Run the `external-data` chunk in ClearanceProject-1-Data.Rmd
load(file = "data-raw/CYP_Sipes2017.Rdata")

@@

@@

system.time(
  temp_lst <- parLapply(clust, temp_lst, cvt_httk_predictions)
)
stopCluster(clust)

@@

fit_tree_rat <- rpart::rpart(Bin ~ ., data = tree_rat)

png(
  filename = paste0(
    Sys.getenv("FIG_DIR"),
    "RestrictiveClearance_Analysis/Rat_Tree.png"
  ),
  width = 4, height = 3, units = "in", bg = "white",
  res = 200
)
rpart.plot::rpart.plot(fit_tree_rat)
dev.off()

preds_tree <- predict(fit_tree_rat, tree_rat, type = "class")
caret::confusionMatrix(preds_tree, tree_rat$Bin, positive = 'Distinguished')

saveRDS(fit_tree_rat, file = "data-raw/RCdata/fit_tree_rat.rds")
